### 2023-07-01 src/latex/AdvPLIC.tex
Update to ratified version 1.0
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

\textbf{%
This chapter is \emph{frozen}, meaning
a functional change is extremely unlikely.
Change will occur only because of some truly critical issue
being identified during public review.%
}
\bigskip

@@

This chapter specifies an \emph{Advanced PLIC} (APLIC) that is not backward
compatible with the earlier {\RISCV} PLIC.
Full conformance to the Advanced Interrupt Architecture requires the
APLIC.
However, a workable system can be built substituting the older PLIC
instead, assuming only wired interrupts to harts, not MSIs.
Chapter~\ref{ch:DuoPLIC} describes a \emph{\mbox{Duo-PLIC\/}} that is
software-configurable to act as either an original {\RISCV} PLIC or an
APLIC.

@@

>>>>

@@

This chapter specifies an \emph{Advanced PLIC} (APLIC) that is not backward
compatible with the earlier {\RISCV} PLIC.
Full conformance to the Advanced Interrupt Architecture requires the
APLIC.
However, a workable system can be built substituting the older PLIC
instead, assuming only wired interrupts to harts, not MSIs.

@@

\begin{commentary}
A draft specification exists for a\/ \emph{\mbox{Duo-PLIC}}
that is software-configurable to act as either
an original {\RISCV} PLIC or an APLIC\@.
However, at this time, it appears unlikely that
the {\RISCV} International Association will ever
ratify the Duo-PLIC specification as a standard.
\end{commentary}

### 2023-07-01 src/latex/IOMMU.tex
Update to ratified version 1.0
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

\textbf{%
This chapter is \emph{frozen} and has already passed public review,
making a functional change at this stage extremely unlikely.%
}
\bigskip

@@

This chapter uses the term \mbox{IOMMU} in a generic sense that encompasses
all translation and transaction processing services required to virtualize
device accesses and is concerned only with how an \mbox{IOMMU} recognizes
and processes MSIs directed to virtual machines. Most other functions and
details of an \mbox{IOMMU} are beyond the scope of this standard, and must
be specified elsewhere.

\begin{commentary}
The {\RISCV} \mbox{IOMMU} Architecture Specification provides a detailed
description of the \mbox{IOMMU} architecture, dividing translation and
transaction processing functionality into blocks such as \mbox{IOMMU},
\mbox{IO Bridge}, etc. and describing how those blocks are integrated
into a system.
\end{commentary}

@@

If a naturally aligned 32-bit write is to guest physical address~$A$
within a page covered by an MRIF-mode PTE, and if the write
data is~$D$ when interpreted in the byte order indicated
by bit~2 of $A$, then the write is processed as follows:
If either $A$\z{[}11:3\z{]} or $D$\z{[}31:11\z{]} is not zero, or if
bit~2 of $A$ is one and big-endian MSIs are not supported, then the
incoming write is accepted but discarded.
Else, the original incoming write is recognized as an MSI and is
replaced by one of the following memory accesses, setting the
interrupt-pending bit that corresponds to the interrupt identity $D$
in the destination MRIF to one:
\begin{enumerate}

\item
An atomic AMOOR operation, if the atomic updates are supported, or

\item
A non-atomic read-modify-write sequence, if non-atomic updates are supported

\end{enumerate}
Once the MRIF update operation is visible to all agents in the system,
the \mbox{11-bit} NID value is zero-extended to 32~bits, and this value
is written to the address NPPN\,\z{<<}\,12 (i.e., physical page number NPPN,
page offset zero) in little-endian byte order.

>>>>

@@

This chapter uses the term \emph{\mbox{IOMMU}}
in a generic sense that encompasses
all translation and transaction processing services required to virtualize
device accesses and is concerned only with how an \mbox{IOMMU} recognizes
and processes MSIs directed to virtual machines. Most other functions and
details of an \mbox{IOMMU} are beyond the scope of this standard, and must
be specified elsewhere.

\begin{commentary}
The {\RISCV} \mbox{IOMMU} Architecture Specification provides a detailed
description of the \mbox{IOMMU} architecture, dividing translation and
transaction processing functionality into blocks such as \mbox{IOMMU},
\mbox{IO Bridge}, etc.\ and describing how those blocks are integrated
into a system.
\end{commentary}

@@

If a naturally aligned 32-bit write is to guest physical address~$A$
within a page covered by an MRIF-mode PTE, and if the write
data is~$D$ when interpreted in the byte order indicated
by bit~2 of $A$, then the write is processed as follows:
If either $A$\z{[}11:3\z{]} or $D$\z{[}31:11\z{]} is not zero, or if
bit~2 of $A$ is one and big-endian MSIs are not supported, then the
incoming write is accepted but discarded.
Else, the original incoming write is recognized as an MSI and is
replaced by one of the following memory accesses, setting the
interrupt-pending bit that corresponds to the interrupt identity $D$
in the destination MRIF to one:
\begin{tightList}
\item
an atomic AMOOR operation, if atomic updates are supported; or
\item
a non-atomic read-modify-write sequence, if atomic updates are not supported.
\end{tightList}
Once the MRIF update operation is visible to all agents in the system,
the \mbox{11-bit} NID value is zero-extended to 32~bits, and this value
is written to the address NPPN\,\z{<<}\,12 (i.e., physical page number NPPN,
page offset zero) in little-endian byte order.

### 2023-07-01 src/latex/MSLevel.tex
Update to ratified version 1.0
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

The AIA does not itself require that detected RAS events trigger
one of the two local interrupts defined for this purpose.
Systems are free to report any or all RAS events
another way, such as by external interrupts routed
through an APLIC or IMSIC, or by custom interrupts.
In all likelihood, the method for reporting a particular RAS event
will depend on where in the system the event is detected.
The AIA defines local interrupt numbers for RAS events so systems
have a standard way to report such events when detected locally at
a hart, without depending solely on external or custom interrupts.

\begin{commentary}
As always, platform standards may further constrain how
a system reports events, whether RAS events or other.
\end{commentary}

>>>>

The AIA does not itself require that detected RAS events trigger
one of the two local interrupts defined for this purpose.
Systems are free to report any or all RAS events
another way, such as by external interrupts routed
through an APLIC or IMSIC, or by custom interrupts.

\begin{commentary}
In all likelihood, the method for reporting a particular RAS event
will depend on where in the system the event is detected.
The AIA defines local interrupt numbers for RAS events so systems
have a standard way to report such events when detected locally at
a hart, without depending solely on external or custom interrupts.

As always, platform standards may further constrain how
a system reports events, whether RAS events or other.
\end{commentary}
